Detection and characterisation of stellar multiplicity from \Gaia\ data

### 2018-06-03 article/ms.tex
words

%\graphicspath{../figures}

@@

\title{Detection of millions of stellar binaries using \Gaia\ data}

@@

\author[0000-0003-0174-0564]{Andrew R. Casey}
\affiliation{School of Physics \& Astronomy, Monash University, Clayton 3800, 
    Victoria, Australia}
\affiliation{Faculty of Information Technology, Monash University, Clayton 3800, 
    Victoria, Australia}

\author{Daniel Foreman-Mackey}
\affiliation{Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}

%\author[0000-0003-2866-9403]{David W. Hogg}
%\affiliation{Center for Cosmology and Particle Physics, Department of Physics,
%    New York University, 726 Broadway, New York, NY 10003, USA}
%\affiliation{Center for Data Science, New York University, 60 Fifth Ave, 
%    New York, NY 10011, USA}
%\affiliation{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 
%    Heidelberg}
%\affiliation{Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}

\begin{abstract}
Most stars are in binary systems. The occurrence rate of multiple order systems 
(e.g., binaries, triples) is relevant to many fields of astrophysics, but the 
effects of stellar multiplicity are commonly ignored in the analysis of most 
stellar and galactic surveys. Here we use \Gaia\ data to identify 
\todo{\NumberOfHigherOrderSystems} higher-order (e.g., binary) systems based on
radial velocity or astrometric jitter, photometry, and for a subset of systems, 
we show that the \emph{lack} of reported radial velocities can be a reliable 
indicator of double-lined spectroscopic binary systems. Our sample includes
\todo{\NumberOfSingleLinedSystems} unresolved binary systems detected solely
from radial velocity or astrometric jitter, which have orbital periods up to 
$\approx$3.5\,yrs. We show that the ratio of jitter in radial velocity and
astrometry can be calibrated to estimate inclination angles (and companion masses)
for thousands of systems. \todo{We comment on stellar multiplicity with stellar
metallicity}. \todo{We comment on systems with stellar remnants (e.g. NS/BHs).}
\end{abstract}

@@

\begin{itemize}
\item the binary fraction is important
\item Gaia provides exquisite astrometry and RV measurements over many years
\item The errors in these quantities are well-characterised
\item When a star is in a binary system (or triary, etc), there will be excess scatter relative to a single star of the primary's properties
\item the excess RV noise (and in DR1 or DR3, excess astrometric noise) can be used to infer the presence of a companion
\item here we do that for lots of sources
\end{itemize}

@@

There are many ways that binaries can be reliably identified using \Gaia\
data, including radial velocity measurements, photometry, and astrometry.
These methods are sensitive to binary systems with different properties.
Here we describe the radial velocity information that forms part of our 
model before describing how photometry and astrometry are incorporated into our
model.

\subsection{SB1 Systems: Single-lined spectroscopic binaries}

Radial velocity measurements are not available for individual epochs in the 
second \Gaia\ data release, but the median and error of radial velocity
measurements are available for 7,224,631 point sources \cite{gaia_rvs}. 
The radial velocity error provided ($\GaiaRVE$; column \texttt{radial\_velocity\_error}) 
is a function of the number of transits $N$ (i.e., the number of observations; 
represented by column \texttt{rv\_nb\_transits}) and the scatter among those $N$ measurements. 
This allows us to calculate the scatter among those measurements, which we will
refer to as the \emph{radial velocity jitter}, for each source
\begin{eqnarray}
\RVJitter = \sqrt{\frac{2N}{\pi}}\GaiaRVE \quad .
\end{eqnarray}

For a single star system observed $N$ times, the source radial velocity
jitter represents either the minimum uncertainty with which \Gaia\ can measure 
radial velocity for a star of that apparent brightness and colour. However for 
stars with stellar companions and no evidence of double lines or binarity in the
cross-correlation-function, so-called SB1 type systems \citep{someone}, the
radial velocity jitter is a function of the orbital parameters and observing
epochs. We assume that the distribution of radial velocity jitter in the second
\Gaia\ data release is a mixture of two populations:

@@

\begin{enumerate}
    \item the RV error is or is not a function of colour
    \item the template used for RV determination does or does not affect the 
          error floor
    \item the jitter in radial velocity measurements for a single star will
          increase as a function of apparent flux, etc.
    \item the spread in the jitter will also increase as a function of these
          properties
\end{enumerate}

Given these assumptions, the model we adopt for the radial velocity jitter is
\begin{eqnarray}
    MODEL
\end{eqnarray}

@@

\subsection{SB2 Systems: Double-lined spectroscopic binaries}

Many sources in the second \Gaia\ data release do not have any radial velocity
measurements reported, despite being bright enough ($G \lesssim 13$) and in a
suitable temperature range for radial velocities to be measured. The principle
reason why radial velocity measurements are \emph{not} reported for these stars
is because the \Gaia/DPAC team have identified the source to be a double-lined
spectroscopic binary (a SB2-type system), in which case it is not sensible to 
report a single radial velocity measurement. For fainter or bluer sources, 
however, the radial velocity may not be reported simply because it is too faint, 
blue, or red.

We calculated the completeness of radial velocity measurements (e.g., the
fraction of sources with reported radial velocities) as a function of all 
available source properties that might affect whether the radial velocity may
be reported or not. This included position ($\alpha$, $\delta$, $l$, $b$),
parallax, proper motions, apparent magnitudes, colours, properties of the radial
velocity templates, stellar parameters ($\teff$, $\radius$, $\luminosity$),
and other properties. The full list of properties investigated is described in
Appendix \todo{X}. In Figure \ref{fig:radial_velocty_completeness} we show the
completeness as a function of just some of these properties, which demonstrate 
that the radial velocity completeness is relatively flat until a star becomes 
too faint (low \rp\ flux), or is either too blue or too red. We adopt 
conservative limits for when the radial velocity completeness starts to drop 
with these properties, and we assume that any source within the following
parameter range

\begin{eqnarray}
    \todo{X} =& \textrm{phot rp mean flux}  = &\todo{X} \\
    \todo{X} =& \textrm{bp}-\textrm{rp}    = &\todo{X}
\end{eqnarray}
\noindent{}is likely to be a double-lined spectroscopic binary (SB2) if no 
radial velocity is reported.

We caution that this deductive inference is potentially far more uncertain (or
contaminated) than the binary probabilities we calculate from the radial
velocity jitter, astrometric excess noise, and photometric colours. Here we are
explicitly assuming that if a source is bright enough and within the right 
colour range, then a radial velocity would be provided \emph{unless} the source
is a spectroscopic binary. In Section \ref{sec:discussion} we detail some of the
immediate limitations or interpretations of this assumption, but here we simply
note that the probability of a source being a double-lined spectroscopic binary
is calculated (or rather, deduced) separately to our calculations of binarity
probability given \emph{measured} quantities like radial velocity, astrometry,
and photometric colours.

\begin{itemize}
    \item completeness
    \item cross-check against SB catalogues -- does it match expectations?
    \item what about close to the plane? 
    \item when and why is this bad?
\end{itemize}

>>>>

@@

\title{Detection and characterisation of stellar multiplicity from \Gaia\ data}

@@

\author[0000-0003-0174-0564]{Andrew R. Casey}
\affiliation{School of Physics \& Astronomy, 
			 Monash University,
			 Wellington Rd, Clayton 3800, Victoria, Australia}
\affiliation{Faculty of Information Technology, 
			 Monash University, 
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author[0000-0002-9328-5652]{Daniel Foreman-Mackey}
\affiliation{Flatiron Institute, 
			 162 Fifth Ave, New York, NY 10010, USA}

\author[0000-0003-2866-9403]{David W. Hogg}
\affiliation{Center for Cosmology and Particle Physics, Department of Physics,
    		 New York University, 
		  	 726 Broadway, New York, NY 10003, USA}
\affiliation{Center for Data Science, 
			 New York University, 60 Fifth Ave, New York, NY 10011, USA}
\affiliation{Max-Planck-Institut f\"ur Astronomie, 
			 K\"onigstuhl 17, D-69117 Heidelberg}
\affiliation{Flatiron Institute,
			 162 Fifth Ave, New York, NY 10010, USA}

\begin{abstract}
Most stars are in binary systems. The occurrence rate and properties of stellar 
multiples (e.g., binaries, triples) is relevant to many fields of astrophysics, 
but the effects of stellar multiplicity are commonly ignored in the analysis of 
most stellar and galactic surveys. Here we use \Gaia\ data to identify 
\todo{\NumberOfHigherOrderSystems} stellar multiple systems from radial velocity
and astrometric jitter, photometry, and intrinsic luminosity. Our sample includes
\todo{\NumberOfSingleLinedSystems} unresolved binary systems detected solely
from radial velocity or astrometric jitter, which have orbital periods up to 
$\approx$3.5\,yrs. We show that the ratio of jitter in radial velocity and
astrometry can be calibrated to estimate inclination angles (and companion masses)
for thousands of systems. \todo{We comment on stellar multiplicity with stellar
metallicity}. \todo{We comment on systems with stellar remnants (e.g. NS/BHs).}
\end{abstract}

@@

Higher order stellar systems describes any stars that have at least one stellar 
companion (e.g., binaries, trinaries). The presence of stellar companions
introduces a number of complications to the inference of stellar and galactic
properties, but these factors are often ignored.

The \Gaia\ space telescope provides exquisite astrometry, photometry, and radial
velocity measurements over many years for million of point stars in our galaxy.
%These data provides an excellent opportunity to self-calibrate the errors in those
%measurements. 
When a point source has an unresolved stellar companion within some range of orbital
and stellar parameters, it is expected that there will be an anomalous excess in 
these measured properties relative to a single star of a similar type. This is
nothing new: astronomers have used radial velocity \citep{firstplanet} and 
astrometric variations \citep{someone} and photometry \citep{someone} to infer the 
presence of stellar and exoplanet companions forever. However, the opportunity
presented by \Gaia\ is to infer the presence of stellar companions for millions
of point sources, across both hemispheres. No ground- or space-based instrument
has ever provided such an opportunity.

In this work we make use of photometry and astrometry (including radial motion)
measurements in the second \Gaia\ data release to infer the presence of stellar
companions for millions of point sources, and provide some basic characterisations
of these systems. In Section \ref{sec:method} we describe our methods, and in
Section \ref{sec:results} we outline our results in context of other stellar
multiplicity surveys. We include comparisons of binary properties between
clusters and the field, as a function of stellar properties (e.g., stellar
metallicity), and highlight some particularly noteworthy systems that would
benefit from immediate follow-up spectroscopic observations.

@@

There are many ways that binaries can be reliably identified using \Gaia\
data, including radial velocity measurements, photometry, and astrometry.
These detection methods are sensitive to binary systems with different 
properties. Here we describe the radial velocity information that forms 
part of our model before describing how photometry and astrometry are 
incorporated into the detection and characterisation of stellar multiples.

\subsection{SB1 Systems: Single-lined spectroscopic binaries}
\label{sec:sb1_methods}

Radial velocity measurements are not available for individual epochs in the 
second \Gaia\ data release \citep{Lindegren:2018,Gaia_RVS}, but the a radial velocity point estimate and error
is available for 7,224,631 sources \cite{gaia_rvs}. 
The radial velocity error provided by \Gaia\ ($\GaiaRVE$; column name \texttt{radial\_velocity\_error}) 
is a function of the number of transits $N$ (i.e., the number of observations; 
represented by column \texttt{rv\_nb\_transits}) and the scatter among those $N$ measurements. 
This allows us to calculate the scatter among those measurements, which we will
refer to as the \emph{radial velocity jitter}, for each source
\begin{eqnarray}
\RVJitter = \sqrt{\frac{2N}{\pi}}\GaiaRVE \quad .
\end{eqnarray}

For a single star (without a significant mass companion) the source radial velocity
jitter represents the minimum uncertainty with which \Gaia\ can measure 
radial velocity for a star of that apparent brightness and colour. However for 
stars with stellar companions and no evidence of double lines or binarity in the
cross-correlation-function, so-called SB1 type systems \citep{someone}, the
radial velocity jitter is a function of the orbital parameters and observation
epochs. We assume that the distribution of radial velocity jitter in the second
\Gaia\ data release is a mixture of two populations:

@@

\begin{itemize}
	\item the radial velocity jitter is a function of the apparent brightness 
		  of the point source
	\item the radial velocity jitter is a function of the apparent $bp-rp$
		  colour of the point source
	\item there is some intrinsic spread to the radial velocity jitter of
		  single stars, and that intrinsic spread also increases with the 
		  apparent brightness and colour of the source
	\item the difference between the true surface gravity $\logg$ and the
		  surface gravity of the radial velocity template used
		  (column \texttt{rv\_template\_logg}) has no impact on the radial
		  velocity jitter
	\item the difference between the true stellar metallicity [Fe/H] and
	 	  the metallicity of the radial velocity template used
		  (column \texttt{rv\_template\_feh}) has no impact on the radial
		  velocity jitter
	\item \todo{state something about teff template}		  
\end{itemize}

Conditioned on these assumptions, we adopt a two-component mixture model for 
the radial velocity jitter: a normal distribution where the mean and variance
are a function of stellar properties, and a uniform background component to
account for stellar multiples. Specifically, the likelihood is

\begin{eqnarray}
%	\mathcal{L} & \equiv & p\left(\{y
    MODEL
\end{eqnarray}

@@

\subsection{SB2 Systems: Double-lined spectroscopic binaries}
\label{sec:sb2_methods}

Many sources in the second \Gaia\ data release do not have any reported radial 
velocity measurement, despite being bright enough ($G \lesssim 13$) and in a
suitable temperature range (e.g., between $X$ and $Y$) for radial velocities to 
be measured \citep{someone}. The principle reason why radial velocity measurements 
are \emph{not} reported for these stars is because the \Gaia/DPAC team have 
identified the source to be a double-lined spectroscopic binary (a SB2-type system), 
either through a composition of two stellar sources present in the spectra, or
from multiple (significant) modes in the cross-correlation function. In these
situations it is not sensible to report a point estimate of the radial velocity
of the point source. For fainter or bluer sources, however, the radial velocity 
may not be reported simply because it is too faint, blue, or red.

We calculated the completeness of radial velocity measurements (e.g., the
fraction of sources with reported radial velocities) as a function of all 
available source properties that might affect whether the radial velocity may
be reported or not. This included position ($\alpha$, $\delta$, $l$, $b$),
parallax, proper motions, apparent magnitudes, colours, properties of the radial
velocity templates, stellar parameters ($\teff$, $\radius$, $\luminosity$),
and other properties. The full list of properties investigated is described in
Appendix \ref{appendix:sb2}. In Figure \ref{fig:radial_velocity_completeness} 
we show the completeness as a function of some pertinent properties, which demonstrate 
that the radial velocity completeness is relatively flat until a star becomes 
too faint (low \rp\ flux), or is either too blue or too red. We adopt 
conservative limits for when the radial velocity completeness starts to drop 
with these properties, and we assume that any source within the following
parameter range

\begin{eqnarray}
    \mathrm{phot\,\,rp\,\,mean\,\,flux} & \in & [10^{4}, 10^8] \nonumber \\
    \mathrm{bp\,\,-rp} & \in & [X, Y] 
    \label{eq:sb2_criteria}
\end{eqnarray}
\noindent{}is likely to be a double-lined spectroscopic binary (SB2) if no 
radial velocity is reported. That is to say that we are explicitly assuming
that if a point source meets the criteria in Equation \ref{eq:sb2_criteria} 
and does not have a radial velocity measurement, then the point source is
an unresolved double-lined spectroscopic binary. In principle we could
make more realistic attempts to model the radial velocity completeness as
a function of stellar properties rather than simply stating ``sources within
this parameter range should have radial velocities unless they are SB2 systems'',
but the radial velocity completeness within our specified range is reasonably
flat.

In Section \ref{sec:discussion} we detail some of the immediate limitations or 
interpretations of this assumption, but here we  note that the probability of a 
source being a double-lined spectroscopic binary is calculated (or rather, 
deduced) separately to our calculations of binarity probability given 
\emph{measured} quantities like radial velocity, astrometry, and photometric 
colours. And as we discuss in Section \ref{sec:discussion}, the main conclusions
of this work do not depend on SB2 systems. Nevertheless, we \emph{strongly} 
caution that our deductive inference of SB2 systems will be far more uncertain 
than the binary probabilities we derive from other information. Our indicator 
whether a star is likely a SB2 system likely suffers from more contamination
and lower completeness compared to the probabilities we derive from other information
(e.g., the radial velocity jitter, astrometric excess noise, and photometric 
colours), and the contamination and completeness likely vary over some
complex (unknown) function of parameter space.

% Figure: radial velocity completeness as a function of pertinent properties
% 		  with the range that we have said is "fully complete"

% Figure: h-r diagram of SB2 binary fraction

% Figure: cross-match against a catalog of SB2 systems,... do we find all of
% 		  what they find from SB2 deductive inference alone? or do we miss
% 		  some SB2s in some parameter range?

% Figure: radial velocity fraction as a function of galactic latitude and
% 		  longitude
